Phylogeographic inference with microtellite and SNPs

### 2019-07-09 code/rmarkdown/article.Rmd
clumpak

The above-mentioned sharp clustering with large-scale SNP data compared to microtellites using non-parametric methods does not reflect when using model-based clustering in the popular software STRUCTURE. This software provides a powerful tool to estimate individual ancestries from genotype data (Pritchard et al. 2000). The modeling of allele frequencies can be more powerful (Nassir et al. 2008), but comes at a high computer cost compared to non-parametric clustering (Porras-Hurtado et al. 2013), and in many cases, methods such as discriminant analysis of principal components claim to perform better (Jombart et al. 2008). Simulations show equivalent power of tens of microtellites and thousands of SNPs to retrieved with high confidence the correct number of cluster (Haasl and Payseur 2011). With empirical data, overall genetic structure with both datasets can vary in the number of most optimal number of clusters (best K) and proportions of individual ancestries (Bradbury et al. 2015; Malenfant et al. 2015; Elbers et al. 2016; Bohling et al. 2019) or be practically identical (Lemopoulos et al. 2019). However, the best K can be very sensitive to demographic processes and can vary substantially depending on the method to estimate it (Janes et al. 2017; Wang 2019). Other issues as uneven sampling and hierarchical population structure can also difficult the inference of the correct population structure (Puechmaille et al. 2016; Wang 2016). Neighbor joining clustering is a complementary way to overcome some limitations of model-based clustering and to recover a reliable evolutionary history of the populations under a wide variaty of scenarios (Kalinowski 2009, 2011). The reconstructions of phylogenetic trees using neighbor joining with microtellite data are often coherent with the phylogeographic structure recovered with SNPs. However, the large homoplasy undermines their power to retrieve phylogenetic relationships between distant populations (Estoup et al. 2002). So, trees from microsatellite data recurrently have little support, specially at deeper nodes, which limits the interpretations of the evolutionary history of the populations (Moore et al. 2014; Candy et al. 2015; Bruneaux et al. 2012; Bradbury et al. 2018), although when several hundreds of loci are used the correct topology is recovered correctly (Takezaki and Nei 2006). The large amount of SNPs and their genome-wide distribution secure a range mutation rates which can provide sufficient information at different evolutionary scales, from recent within species processes to interspecies phylogenies (eg. Defaveri et al. 2013; Petersen et al. 2013).

@@

@@

## 3.1 Model-based clustering
Convergence of the chains measured by the GR diagnostic varied between species and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).

* result 3: genetic diversity

@@

* Key result 2

>>>>

The above-mentioned sharp clustering with large-scale SNP data compared to microtellites using non-parametric methods does not reflect when using model-based clustering in the popular software STRUCTURE. This software provides a powerful tool to estimate individual ancestries from genotype data (Pritchard et al. 2000). The modeling of allele frequencies can be more powerful (Nassir et al. 2008), but comes at a high computer cost compared to non-parametric clustering (Porras-Hurtado et al. 2013), and in many cases, methods such as discriminant analysis of principal components claim to perform better (Jombart et al. 2008). Simulations show equivalent power of tens of microtellites and thousands of SNPs to retrieved with high confidence the correct number of cluster (Haasl and Payseur 2011). With empirical data, overall genetic structure with both datasets can vary in the number of most optimal number of clusters (best K) and proportions of individual ancestries (Bradbury et al. 2015; Malenfant et al. 2015; Elbers et al. 2016; Bohling et al. 2019) or be practically identical (Lemopoulos et al. 2019). However, the best K can be very sensitive to demographic processes and can vary substantially depending on the method to estimate it (Janes et al. 2017; Wang 2019). Other issues as uneven sampling and hierarchical population structure can also difficult the inference of the correct population structure (Puechmaille et al. 2016; Wang 2016). Neighbor joining clustering is a complementary way to overcome some limitations of model-based clustering and to recover a reliable evolutionary history of the populations under a wide variety of scenarios (Kalinowski 2009, 2011). The reconstruction of phylogenetic trees using neighbor joining with microtellite data are often coherent with the phylogeographic structure recovered with SNPs. However, the large homoplasy undermines their power to retrieve phylogenetic relationships between distant populations (Estoup et al. 2002). So, trees from microsatellite data recurrently have little support, specially at deeper nodes, which limits the interpretations of the evolutionary history of the populations (Moore et al. 2014; Candy et al. 2015; Bruneaux et al. 2012; Bradbury et al. 2018), although when several hundreds of loci are used the correct topology is recovered correctly (Takezaki and Nei 2008). The large amount of SNPs and their genome-wide distribution secure a range mutation rates which can provide sufficient information at different evolutionary scales, from recent within species processes to interspecies phylogenies (eg. Defaveri et al. 2013; Petersen et al. 2013).

@@

## 2.4 Phylogenetic structure

We evaluated the phylogenetic relationships between the populations using the neighbor-joining (NJ) algorithm (Saitou and Nei 1987). We discarded using model-based phylogenetic inference for the implicit difficulties of fitting evolutionary and demographic models for intra-species trees (), and for the possibilities that offer comparing the 2 marker types (SNPs and microtellites) under a similar tree-reconstruction method. The application of NJ relies on a distance matrix computed from the genotypes. For the microtellite datasets we computed these matrices using Da distance (Nei et al. 1983) with *genet.dist* function from *hierfstat* 0.04-22 package (Goudet and Jombart 2015) in R. Da distances have been shown to yield more reliable topologies to other distances (Takezaki and Nei 1996, 2008). For the SNP genotypes, we computed euclidean distances with *stats::dist* function in R. Then NJ trees were reconstructed with *njs* function from *ape* 5.0 (Paradis and Schliep 2018) for each dataset. We computed the bootstrap support with a modified approach from *ape::boot.phylo* to accommodate non-standard input data. First, we resampled with replacement loci in the genotype matrices, from which we recomputed NJ trees. These trees were rooted using midpoint rooting with function *midpoint.root* from *phytools* 0.6-60 (Revell 2012) in R. And finally, bootstrap values were computed by counting the partitions of the original trees present in 1000 bootstrapped trees.
*Marker type and bootstrap support*. For each tree, branch lengths were normalized by dividing edge lengths by the length between the most distant tips in the tree. Then, we used several functions in *treeman* 1.1.3 package (Bennett 2018) in R, to extract 2 metrics from each node in the trees: 1) the distance from each node to the root (Dnr), and 2) the distance from each node to its parent node (Dnp). ??For the nodes in each tree we assessed the relation of Dnr and Dnp against their corresponding bootstrap support values fitting generalized linear models (GMLs) using a logit link function, with *stats::glm* function in R. *Actually the fit is really bad. See script 99_04_model_fitting.R and ask Jeff*??

*Number of markers and bootstrap support*.

*MAF and bootstrap support*.

@@

## 3.1 Model-based clustering
*Convergence*. Convergence of the chains measured by the GR diagnostic varied between species and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).
*Optimal K*. The most optimal number of populations was evaluated with 3 metrics. Pr[X|K] increased steadily from K = 2 to K = 8 in all datasets (Figure 2 A-C). Evanno's ΔK peaked at K = 2 in all SNP datasets, followed by K = 3 (Figure 2 D-F). However, in the microsatellite dataset from *H. molleri*, ΔK peaked at K = 4 and K = 6 (Figure 2D). The predominant K = 2 in the SNP datasets were not in agreement with results from the Parsimony index (Figure 2 G-I), except for the microtellite data in *H. molleri*, in which PI and ΔK peaked at K = 4 (Figure 2G). For SNPs in *H. molleri*, the highest values of the PI were at K = 4 and K = 5, contrary to the K = 2 from ΔK (Figure 2G). In *P. cultripes*, the PI peaked at K = 4 for both markers, and then sunk in the SNP data, whereas for  microtellites the PI reached its maxima at K = 6 - K = 8 (Figure 2H). In the subsets of SNPs from *P. cultripes* the PI increased gradually up to K = 4. After that, the distinct marker sets showed inconsistent trends (Figure 2I).

## 3.2 Phylogenetic structure

*Coherence between model-based and NJ clustering*.

*Marker type and bootstrap support*.

*Number of markers and bootstrap support*.

*MAF and bootstrap support*.

@@

* Key result 2
The little bootstrap support we describe in phylogenetic trees from microtellite data is not surprising, given that often several tens of microsatellites or more are needed to recover the correct topology of trees (Nei et al. 1983; Takezaki and Nei 1996)
